A Fast kNN Algorithm Using Multiple Space-Filling Curves

### 2022-03-19 2022/Sensors/sensors.tex
fix some mistakes and references

Currently, machine learning (ML) methods are being used successfully to solve a wide range of problems in various application areas. One example of a class of problems where ML has demonstrated its effectiveness are the tasks of identifying the main properties of the phenomena, which are characterized by their stochastic nature or the presence of hidden parameters \cite{Golovenkin2020, Gonoskov2019, Kastalskiy2021SocialSD}.

@@

Among the undoubted merits of kNN is its "explainability", since the decision about whether a test object belongs to one class or another is clearly explained by similar properties of the test object and its nearest (in some metric) neighbours.
Along with the study of the method’s theoretical properties, the issue of its effective implementation is also investigated. Thus, direct implementation of the method by using the  "brute-force" approach has computational complexity $O(n)$, where $n$ is the number of objects. Fast algorithms, which have computational complexity $O(log(n))$, use complex data structures of kd-tree type [kdtree-1,2] \cite{Bentley1975MultidimensionalBS}.

@@

In this study, the classification problem will be understood as the problem of assigning an object $\omega$  to one of the predetermined classes based on its features. We will assume that each of the objects $\omega$ is represented as a feature vector

\begin{equation}
y = (y_1, y_2, ..., y_N) \in R^N,
\end{equation}

where the value of the $j$-th vector coordinate corresponds to the numerical value of the $j$-th feature of the object $\omega$.

@@

The general scheme of object classification using kNN can be formulated as follows

\begin{itemize}
\item calculate the distance from the object being classified to each of the objects in the training set,
\item select k objects of the training set with the minimum distance to them,
\item the class of the object being classified is the class most often found among the k nearest neighbours.
\end{itemize}

@@

The main drawback of the brute force method is the unacceptable running time and the rapid growth of the computation volume as the number of objects increases. This method can be applied when the number of objects is relatively small and the dimension of $y$ is small, and when the "curse of dimensionality" is not yet in full effect.

@@

As it was mentioned, the computational cost of searching for nearest objects increases with increasing dimensionality, both in the case of exhaustive search for all distances and when using tree structures. One mechanism for speeding up the search is dimensionality reduction, by reducing the problem of searching in a multidimensional space to the search in a one-dimensional space (over an interval). This is possible through the use of space-filling curves (Peano-Hilbert curves) that fill the multidimensional space. Such curves are used, for example, in multidimensional global optimization problems [ссылки], in image processing [ссылки], etc.

@@

Note that in practical classification tasks, different features can have different scales, which can significantly distort the real distance between objects. Therefore it will be assumed that prior to applying kNN data scaling has been performed

\begin{equation}
y_j \equiv \frac{y_j - y_{min}} {y_{max} - y_{min}} - \frac{1} {2}.
\end{equation}

Thus, the variation domain for all feature values will form a unit hypercube

\begin{equation}
D = \{ y \in R^N: -1/2 \leq y_j \leq 1/2,  1 \leq j \leq N \}.
\end{equation}

The $1/2$ hypercube offset  has been made for the convenience of further labelling.

@@

Note that the centre  $y^c(x)$ of the $M$-th partition subcube containing point $y(x)$ can be regarded as an approximation to $y(x)$ with the accuracy of $\varepsilon = 2^{-(M+1)}$, i.e.

\begin{equation}
\max_{1 \leq j \leq N} y_j^c(x)-y_j(x) \leq \varepsilon = 2^{-(M+1)}.
\end{equation}

@@

Let $\omega_i, 1 \leq i \leq S$, be the $i$-th object from the set of objects correlated with class $C_j$. Previously, a feature vector $y = (y_1, y_2, ..., y_N) \in R^N$, corresponding to each object under consideration was defined, i.e.

\begin{equation}
\omega_i \leftrightarrow y = (y_{i1}, y_{i2}, ..., y_{iN})
\end{equation}

Using the Peano curve, each object is assigned the value of $x_i$, i.e.

\begin{equation}
\omega_i \leftrightarrow x_i, \; y(x_i) = (y_{i1}, y_{i2}, ..., y_{iN}).
\end{equation}

Consider a test object $\omega^*$, to which the value $x^*$ is assigned. The degree of proximity of the test object $\omega^*$ to one of the objects $\omega_i$ is determined based on the distance on the one-dimensional scale $d_i = |x_i - x^* |$.
Thus, the closest object to the test object has the following property:

\begin{equation}
d_{min} = \min \{ d_i: 1 \leq i \leq S \}.
\end{equation}

The operation of the algorithm to determine the nearest neighbour of the test object involves two phases: data preparation and the search itself. Data preparation consists of calculating the correspondence $\omega_i \rightarrow x_i$ for each input object, including the test object, and ordering the objects by the value of $x$.

\begin{equation}
x^1(\omega^1) \leq x^2(\omega^2) \leq ... \leq x^S(\omega^S).
\end{equation}

@@

This property makes dimensionality reduction using a single Peano curve practically inapplicable. To overcome this drawback, various approaches have been proposed, for example, the simultaneous use of two or more space-filling curves of different types [Scott1999], data shifting \cite{Liao2001}. In the following section, we propose a constructive way to preserve some information on object proximity during dimensionality reduction, based on the use of multiple same-type evolvents.

@@

Consider a family of Peano curves

\begin{equation}
Y_L(x)=\{ y^0(x), y^1(x), ..., y^L(x) \}
\end{equation} 
instead of a single curve $y(x)$. We will construct the family of curves as follows. Let us introduce a family of hypercubes

\begin{equation}\label{hypercubes}
D_i = \{ y \in R^N: -2^{-1} \leq y_i + 2^{-l} \leq 3 \cdot 2^{-1},  1 \leq i \leq N \}, \; 0 \leq l \leq L,
\end{equation} 
where the hypercube $D_{l+1}$ is obtained by "shifting" the hypercube $D_l$ along the main diagonal by step $2^{-l}$ for each coordinate.

Let a Peano curve type mapping $y^0(x)$ map the interval $[0,1]$ onto the hypercube $D_0$ of (ссылка), i.e.

\begin{equation}\label{hypercubeD0}
D_0 = \{y^0(x): x \in [0, 1] \}.
\end{equation}

Any subcube of the $M$-th partition of the hypercube $D_0$ will have an edge of length $2^{-(M-1)}$ and will be denoted by $D_0(M, v)$, where $v$ is the left boundary point of the subinterval $d(M, v)$ corresponding to this subcube. Then the evolvents $y^l(x)=\{y_1^l(x), ..., y_N^l(x)\}$, whose coordinates are determined by the conditions

\begin{equation}\label{maps}
y_i^l(x) = y_i^{l-1}(x) + 2^{-l}, \; 1 \leq i \leq N, \; 1 \leq l \leq L,
\end{equation} 
map the interval $[0,1]$ onto the corresponding hypercubes  $D_l, \; 1 \leq l \leq L$ (the broken line in the figure shows the image of the interval $[0,1]$, which is obtained by using the evolvent $y^0(x), x \in [0,1]$).

@@

It follows from formulae (\ref{hypercubeD0}) and (\ref{maps}) that if an interval $d(M,v)$ is mapped onto the subcube $D_0(M,v)$, then there exists a family of subcubes

\begin{equation}
D_l(M,v) = y^l(d(M,v)), 1 \leq l \leq L,
\end{equation} 
connected with the corresponding subintervals $d_l(M,v_l) \subset [0,1]$, where $d(M,v) = d_0(M,v_0), v_0 = v$, such that

\begin{equation}
D_l(M,v) = y^l(d_l(M,v)), 1 \leq l \leq L,
\end{equation}

Since the hypercube $D$ of (ссылка) belongs to the general part of the family of hypercubes (\ref{hypercubes}) (the boundary of the hypercube $D$ is highlighted in Fig \ref{fig2}), then, by introducing an additional function

\begin{equation}
g_0(y) = \max \{ |y_i|-2^{-1}: \; 1 \leq i \leq N \},
\end{equation} 
the original hypercube $D$ can be represented as

\begin{equation}
D = \{y^l(x): x \in [0,1], \; g_0(y^l(x)) \leq 0 \}, \; 0 \leq l \leq L,
\end{equation} 
i.e. $g_0(y) \leq 0$, if $y \in D$, and $g_0(y) > 0$ if otherwise. Hence, any point $y \in D$ has its preimage $x^l \in [0,1]$ at every mapping $y^l(x), \; 0 \leq l \leq L$, i.e.

\begin{equation}
y = y^l(x^l), \; x^l \in [0, 1], \; 0 \leq l \leq L.
\end{equation}

@@

\begin{equation}
\begin{array}
\: y^* = y^l(x^*), y^\prime = y^l(x^\prime), y^{\prime\prime} = y^l(x^{\prime\prime}), \\
\max \{ |x^\prime - x^*|, |x^{\prime\prime} - x^*|, |x^\prime - x^{\prime\prime}| \} \leq 2^{-pN}.
\end{array}
\end{equation}

@@

The input data of the algorithm also includes the feature vector $y^*$ corresponding to the test object $\omega^*$.
\\[12pt]
\underline{Procedure A}

\begin{itemize}
\item For each scale, search for $K$ objects closest to $\omega^*$. The result of the search is a set of vectors of  the nearest objects on each scale.
\item The vectors of the objects found are inserted into an RB-tree according to the distance criterion $d_{min}$.
\item After processing all the scales, the first $K$ objects are extracted from the resulting tree and they are considered to be the search result.
\item The number of objects belonging to each class is counted and the decision is made whether   the test object belongs to a particular class.
\end{itemize}
\\[12pt]
\underline{Procedure B}

\begin{itemize}
\item For each scale, search for the closest object to $ω^*$; the result is a set of distances between the test object and the closest object for each scale.
\item The object with minimal distance on all scales is selected.
\item This object is saved in a special vector and excluded from the set of objects.
\item The whole procedure is repeated $K$ times, resulting in a vector of K nearest neighbours.
\item The number of objects belonging to each class is counted and the decision is made whether   the test object belongs to a particular class.
\item If re-classification is required, the procedure is called to recover the previously excluded objects and search for neighbours.
\end{itemize}

>>>>

Currently, machine learning (ML) methods are being used successfully to solve a wide range of problems in various application areas. One example of a class of problems where ML has demonstrated its effectiveness are the tasks of identifying the main properties of the phenomena, which are characterized by their stochastic nature or the presence of hidden parameters \cite{Golovenkin2020, Gonoskov2019, Kastalskiy2021}.

@@

Among the undoubted merits of kNN is its ``explainability'', since the decision about whether a test object belongs to one class or another is clearly explained by similar properties of the test object and its nearest (in some metric) neighbours.
Along with the study of the method's theoretical properties, the issue of its effective implementation is also investigated. Thus, direct implementation of the method by using the  ``brute-force'' approach has computational complexity $O(n)$, where $n$ is the number of objects. Fast algorithms, which have computational complexity $O(log(n))$, use complex data structures of kd-tree type [kdtree-1,2] \cite{Bentley1975MultidimensionalBS}.

@@

In this study, the classification problem will be understood as the problem of assigning an object $\omega$  to one of the predetermined classes based on its features. We will assume that each of the objects $\omega$ is represented as a feature vector 
\begin{equation}
y = (y_1, y_2, ..., y_N) \in R^N,
\end{equation} 
where the value of the $j$-th vector coordinate corresponds to the numerical value of the $j$-th feature of the object $\omega$.

@@

The general scheme of object classification using kNN can be formulated as follows:
\begin{itemize}
\item calculate the distance from the object being classified to each of the objects in the training set;
\item select k objects of the training set with the minimum distance to them;
\item the class of the object being classified is the class most often found among the $k$ nearest neighbours.
\end{itemize}

@@

The main drawback of the brute force method is the unacceptable running time and the rapid growth of the computation volume as the number of objects increases. This method can be applied when the number of objects is relatively small and the dimension of $y$ is small, and when the ``curse of dimensionality'' is not yet in full effect.

@@

As it was mentioned, the computational cost of searching for nearest objects increases with increasing dimensionality, both in the case of exhaustive search for all distances and when using tree structures. One mechanism for speeding up the search is dimensionality reduction, by reducing the problem of searching in a multidimensional space to the search in a one-dimensional space (over an interval). This is possible through the use of space-filling curves (Peano-Hilbert curves) that fill the multidimensional space. Such curves are used, for example, in multidimensional global optimization problems \cite{Strongin2018,Gergel2021}, in image processing \cite{Liang2008,Costa2012,Herrero2015}, etc.

@@

Note that in practical classification tasks, different features can have different scales, which can significantly distort the real distance between objects. Therefore it will be assumed that prior to applying kNN data scaling has been performed 
\begin{equation}
y_j \equiv \frac{y_j - y_{min}} {y_{max} - y_{min}} - \frac{1} {2}.
\end{equation}

Thus, the variation domain for all feature values will form a unit hypercube
\begin{equation}\label{D}
D = \{ y \in R^N: -1/2 \leq y_j \leq 1/2,  1 \leq j \leq N \}.
\end{equation} 
The $1/2$ hypercube offset  has been made for the convenience of further labelling.

@@

Note that the centre  $y^c(x)$ of the $M$-th partition subcube containing point $y(x)$ can be regarded as an approximation to $y(x)$ with the accuracy of $\varepsilon = 2^{-(M+1)}$, i.e.
\begin{equation}
\max_{1 \leq j \leq N} y_j^c(x)-y_j(x) \leq \varepsilon = 2^{-(M+1)}.
\end{equation}

@@

Let $\omega_i, 1 \leq i \leq S$, be the $i$-th object from the set of objects correlated with class $C_j$. Previously, a feature vector $y = (y_1, y_2, ..., y_N) \in R^N$, corresponding to each object under consideration was defined, i.e. 
\begin{equation}
\omega_i \leftrightarrow y = (y_{i1}, y_{i2}, ..., y_{iN})
\end{equation}

Using the Peano curve, each object is assigned the value of $x_i$, i.e.
\begin{equation}
\omega_i \leftrightarrow x_i, \; y(x_i) = (y_{i1}, y_{i2}, ..., y_{iN}).
\end{equation}

Consider a test object $\omega^*$, to which the value $x^*$ is assigned. The degree of proximity of the test object $\omega^*$ to one of the objects $\omega_i$ is determined based on the distance on the one-dimensional scale $d_i = |x_i - x^* |$.
Thus, the closest object to the test object has the following property:
\begin{equation}
d_{min} = \min \{ d_i: 1 \leq i \leq S \}.
\end{equation}

The operation of the algorithm to determine the nearest neighbour of the test object involves two phases: data preparation and the search itself. Data preparation consists of calculating the correspondence $\omega_i \rightarrow x_i$ for each input object, including the test object, and ordering the objects by the value of $x$.
\begin{equation}
x^1(\omega^1) \leq x^2(\omega^2) \leq ... \leq x^S(\omega^S).
\end{equation}

@@

This property makes dimensionality reduction using a single Peano curve practically inapplicable. To overcome this drawback, various approaches have been proposed, for example, the simultaneous use of two or more space-filling curves of different types \cite{Scott1999}, data shifting \cite{Liao2001}. In the following section, we propose a constructive way to preserve some information on object proximity during dimensionality reduction, based on the use of multiple same-type evolvents.

@@

Consider a family of Peano curves
\begin{equation}
Y_L(x)=\{ y^0(x), y^1(x), ..., y^L(x) \}
\end{equation} 
instead of a single curve $y(x)$. We will construct the family of curves as follows. Let us introduce a family of hypercubes
\begin{equation}\label{hypercubes}
D_i = \{ y \in R^N: -2^{-1} \leq y_i + 2^{-l} \leq 3 \cdot 2^{-1},  1 \leq i \leq N \}, \; 0 \leq l \leq L,
\end{equation} 
where the hypercube $D_{l+1}$ is obtained by "shifting" the hypercube $D_l$ along the main diagonal by step $2^{-l}$ for each coordinate.

Let a Peano curve type mapping $y^0(x)$ map the interval $[0,1]$ onto the hypercube $D_0$ of (\ref{hypercubes}), i.e.
\begin{equation}\label{hypercubeD0}
D_0 = \{y^0(x): x \in [0, 1] \}.
\end{equation}

Any subcube of the $M$-th partition of the hypercube $D_0$ will have an edge of length $2^{-(M-1)}$ and will be denoted by $D_0(M, v)$, where $v$ is the left boundary point of the subinterval $d(M, v)$ corresponding to this subcube. Then the evolvents $y^l(x)=\{y_1^l(x), ..., y_N^l(x)\}$, whose coordinates are determined by the conditions
\begin{equation}\label{maps}
y_i^l(x) = y_i^{l-1}(x) + 2^{-l}, \; 1 \leq i \leq N, \; 1 \leq l \leq L,
\end{equation} 
map the interval $[0,1]$ onto the corresponding hypercubes  $D_l, \; 1 \leq l \leq L$ (the broken line in the figure shows the image of the interval $[0,1]$, which is obtained by using the evolvent $y^0(x), x \in [0,1]$).

@@

It follows from formulae (\ref{hypercubeD0}) and (\ref{maps}) that if an interval $d(M,v)$ is mapped onto the subcube $D_0(M,v)$, then there exists a family of subcubes
\begin{equation}
D_l(M,v) = y^l(d(M,v)), 1 \leq l \leq L,
\end{equation} 
connected with the corresponding subintervals $d_l(M,v_l) \subset [0,1]$, where $d(M,v) = d_0(M,v_0), v_0 = v$, such that 
\begin{equation}
D_l(M,v) = y^l(d_l(M,v)), 1 \leq l \leq L,
\end{equation}

Since the hypercube $D$ from (\ref{D}) belongs to the common part of the family of hypercubes (\ref{hypercubes}) (the boundary of the hypercube $D$ is highlighted in Fig.~\ref{fig2}), then, by introducing an additional function 
\begin{equation}
g_0(y) = \max \{ |y_i|-2^{-1}: \; 1 \leq i \leq N \},
\end{equation} 
the original hypercube $D$ can be represented as
\begin{equation}
D = \{y^l(x): x \in [0,1], \; g_0(y^l(x)) \leq 0 \}, \; 0 \leq l \leq L,
\end{equation} 
i.e. $g_0(y) \leq 0$, if $y \in D$, and $g_0(y) > 0$ if otherwise. Hence, any point $y \in D$ has its preimage $x^l \in [0,1]$ at every mapping $y^l(x), \; 0 \leq l \leq L$, i.e.
\begin{equation}
y = y^l(x^l), \; x^l \in [0, 1], \; 0 \leq l \leq L.
\end{equation}

@@

\begin{equation}
 y^* = y^l(x^*), y^\prime = y^l(x^\prime), y^{\prime\prime} = y^l(x^{\prime\prime}), \\
\max \{ |x^\prime - x^*|, |x^{\prime\prime} - x^*|, |x^\prime - x^{\prime\prime}| \} \leq 2^{-pN}.
\end{equation}

@@

The input data of the algorithm also includes the feature vector $y^*$ corresponding to the test object $\omega^*$.

%\\[12pt]
\underline{Procedure A}

\begin{itemize}
\item For each scale, search for $K$ objects closest to $\omega^*$. The result of the search is a set of vectors of  the nearest objects on each scale.
\item The vectors of the objects found are inserted into an RB-tree according to the distance criterion $d_{min}$.
\item After processing all the scales, the first $K$ objects are extracted from the resulting tree and they are considered to be the search result.
\item The number of objects belonging to each class is counted and the decision is made whether   the test object belongs to a particular class.
\end{itemize}

%\\[12pt]
\underline{Procedure B}

\begin{itemize}
\item For each scale, search for the closest object to $\omega^*$; the result is a set of distances between the test object and the closest object for each scale.
\item The object with minimal distance on all scales is selected.
\item This object is saved in a special vector and excluded from the set of objects.
\item The whole procedure is repeated $K$ times, resulting in a vector of K nearest neighbours.
\item The number of objects belonging to each class is counted and the decision is made whether   the test object belongs to a particular class.
\item If re-classification is required, the procedure is called to recover the previously excluded objects and search for neighbours.
\end{itemize}
